---
title: "R-Ladies NL Book-Club"
subtitle: "Advanced R: Functions (Chapter 6)"  
author: "Josephine Daub"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: ["rladies", "rladies", "rladies-fonts"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

```{r xaringan-themer, include=FALSE}

library(xaringanthemer)
library(pagedown)

style_mono_accent(
  header_font_google = google_font("Lora", "500"),
  text_font_google   = google_font("Lato", "400", "400i")
)
```

### Welcome R-Ladies Netherlands Book-Club!

.left[
- R-Ladies is a global organization to promote gender diversity in the R 
community via meetups, mentorship in a safe and inclusive environment. 
]

<br>
--

.left[
- **R-Ladies Netherlands Book-Club** is a collaborative effort between 
RLadies-NL chapters in Nijmengen, Rotterdam, Den Bosch, Amsterdam, Utrecht.
]

--

- We meet every **2 weeks** to go through one of the chapters of Hadley Wickam _Advanced R_, and run through  exercises to put the concepts into practice.  

---

### Today's Session!

<br>

- Starts with a  30-45 min presentation

- Breakout session - we **split** into breakout rooms to practice exercises.

--
<br>
<br>

- Please use the **HackMD** (shared in email and in chat) to present yourself, ask overarching questions, and to find your break out room.

- Use the **chat** to participate in the discussion during the presentation and your breakout session. 

- The Bookclub github repository has also been made available. 

--
<br>
<br>

- Any questions?

---

### Resources

- Solutions to the exercises from _Advanced R_ can be found in the [Advanced R Solutions Book](https://advanced-r-solutions.rbind.io/index.html)  

- The R4DS book club repo has a Q&A section  : [github.com/r4ds/bookclub-Advanced_R](https://github.com/r4ds/bookclub-Advanced_R)

--
<br>

- We are always looking for new speakers! If you are interested, please sign up to present a chapter at https://rladiesnl.github.io/book_club/

---

name: title
class: center, middle

## Functions ##

---

## Outline

The outline for today is: 
--
<br><br>
- Function fundamentals

- Function composition

- Lexical scoping

- Lazy evaluation

- ... (dot-dot-dot)

- Exiting a function

- Function forms
--
<br><br>

**Breakout Sessions**


---

### Let's get to it! 

---

### Function fundamentals

--

In R, a **function** has three parts:
  1. `formals()`: list of arguments that control how you call the function
  2. `body()`: the code inside the function
  3. `environment()`: data structure that determines how the function finds the values associated with the names

--

The formals and body are specified **explicitly** when creating a function
<br>
The environment is specified **implicitly**, based on *where* you define a function

---
### Example

```{r}
f02 <- function(x, y) {
  # A comment
  x + y
}

formals(f02)

body(f02)

environment(f02)

```
---
### Function fundamentals

Functions can have additional `attributes()`, such as `srcref` (source reference):

```{r}
attr(f02, "srcref")
```

--
<br><br>
Exception: **primitive functions** call C code directly:

```{r}
sum
`[`
```
---

### Function fundamentals

Functions are **objects**
--
<br><br>
Create a function object (with `function`) and bind it to a name with `<-`:

```{r eval=FALSE}
f01 <- function(x) {
  sin(1 / x ^ 2)
}
```
--
<br>
Or: you can choose not to give a name to get a **anonymous function**:

```{r eval=FALSE}
lapply(mtcars, function(x) length(unique(x)))
Filter(function(x) !is.numeric(x), mtcars)
integrate(function(x) sin(x) ^ 2, 0, pi)
```
--
<br>
Or: put functions in a **list**:

```{r eval=FALSE}
funs <- list(
  half = function(x) x / 2,
  double = function(x) x * 2
)

funs$double(10)
```
---

### Function composition

--

How to compose **multiple** function calls?

--

**Nest** the function calls

```{r eval=FALSE}
sqrt(mean(x))
```

--

Save **intermediate results**

```{r eval=FALSE}
y<-mean(x)
sqrt(y))
```

--

Use **magrittr** package to use pipe (`%>%`)

```{r eval=FALSE}
library(magrittr)

x %>%
  mean() %>%
  sqrt()
```
---
### Lexical scoping

--

**Scoping**: finding the value associated with a name

--

What will `g01()` return?


```{r eval=FALSE}
x <- 10
g01 <- function() {
  x <- 20
  x
}

g01()
```

--

**R uses lexical scoping**: look up values of names based on how a function is defined, not how it is called 

--

**Four rules**:
1. Name masking
2. Functions versus variables
3. A fresh start
4. Dynamic lookup

---

### Lexical scoping

**Name masking**: names defined inside a function mask names defined outside a function

--

```{r eval=TRUE}
x <- 10
y <- 20
g02 <- function() {
  x <- 1
  y <- 2
  c(x, y)
}
g02()
```

--

If a name is not defined inside a function, R looks **one level up**:

```{r eval=TRUE}
x <- 2
g03 <- function() {
  y <- 1
  c(x, y)
}
g03()
```

---

### Lexical scoping

**Name masking**: names defined inside a function mask names defined outside a function

The same rules apply if a **function is defined inside another function**:


```{r eval=TRUE}
x <- 1
g04 <- function() {
  y <- 2
  i <- function() {
    z <- 3
    c(x, y, z)
  }
  i()
}
g04()
```

---

### Lexical scoping

**Functions vs variables**

--

Scoping rules also apply to functions:

```{r}
g07 <- function(x) x + 1
g08 <- function() {
  g07 <- function(x) x + 100
  g07(10)
}
g08()
```

--

But what happens when a **function** and a **non-function** share the same name?

--

```{r}
g09 <- function(x) x + 100
g10 <- function() {
  g09 <- 10
  g09(g09)
}
g10()
```
---

### Lexical scoping

**A fresh start**

--

What happens to values between invocations of a function?

```{r}
g11 <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  a
}

g11()
```

What happens if you call `g11()` again?

```{r}
g11()
```

---

### Lexical scoping

**Dynamic lookup**

--

R looks for values when the function is **run**, not when the function is **created**

--

```{r}
g12 <- function() x + 1
x <- 15
g12()

x <- 20
g12()
```

--

<br>

Use `codetools::findGlobals()` to list all unbound symbols within a function:

```{r}
codetools::findGlobals(g12)
```

---

### Lazy evaluation

--

Arguments are **lazily evaluated**: they’re only evaluated if accessed

--

```{r}
h01 <- function(x) {
  10
}
h01(stop("This is an error!"))
```

--

<br>

Lazy evaluation is powered by a data structure called a **promise** (see Chapter 20)

<br>
A promise has **three components**:
1. An **expression**
2. An **environment** where the expression should be evaluated
3. A **value**, computed and cached the first time when a promise is accessed in the specified environment


---

### Lazy evaluation

Example:

```{r}
y <- 10
h02 <- function(x) {
  y <- 100
  x + 1
}

h02(y)
```

--

When doing **assignment inside a function call**, the variable is bound **outside** of the function, not inside of it:

```{r}
h02(y <- 1000)
y
```

---
### Lazy evaluation

--

**Default arguments** can be defined in terms of other arguments or variables defined later in the function:

```{r}
h04 <- function(x = 1, y = x * 2, z = a + b) {
  a <- 10
  b <- 100
  
  c(x, y, z)
}

h04()
```

--

**Not recommended!**

---

### Lazy evaluation

**Missing values**

--

You can use `missing()` to determine if an argument’s value comes from the **user** or from a **default**:

```{r}
h06 <- function(x = 10) {
  list(missing(x), x)
}

str(h06())

str(h06(10))
```

---

### Lazy evaluation

**Missing values**

You can use `missing()` to determine if an argument’s value comes from the **user** or from a **default**:

<br>

**Recommendation**: use `missing()` sparingly, instead use  `NULL` to indicate that an argument is not required but can be supplied

--
```{r eval=TRUE}
args(sample)
```


```{r eval=FALSE}
sample <- function(x, size = NULL, replace = FALSE, prob = NULL) {
  if (is.null(size)) {
    size <- length(x)
  }
  
  x[sample.int(length(x), size, replace = replace, prob = prob)]
}
```

---

### ... (dot-dot-dot)

--

Functions can have a special argument: `...`

--

You can use `...` to pass additional arguments on to another function:

```{r eval=TRUE}
i01 <- function(y, z) {
  list(y = y, z = z)
}

i02 <- function(x, ...) {
  i01(...)
}

str(i02(x = 1, y = 2, z = 3))
```

---
### ... (dot-dot-dot)

Functions can have a special argument: `...`

You can use `...` to pass additional arguments on to another function:

```{r eval=TRUE}
args(lapply)

x <- list(c(1, 3, NA), c(4, NA, 6))
str(lapply(x, mean, na.rm = TRUE))
```


---
### ... (dot-dot-dot)

Functions can have a special argument: `...`

You can use `...` to pass additional arguments on to another function

<br>

**Downsides**:

When you use it to pass arguments to another function, **carefully explain** where those arguments go

--

A misspelled argument will not raise an error

```{r eval=TRUE}
sum(1, 2, NA, na_rm = TRUE)
```

---
### Exiting a function

--

Two ways that a function can return a value:

**Implicitly**, the last evaluated expression is the return value:

```{r eval=TRUE}
j01 <- function(x) {
  if (x < 10) {
    0
  } else {
    10
  }
}
j01(5)
```

--

**Explicitly**, by calling `return()`:

```{r eval=TRUE}
j02 <- function(x) {
  if (x < 10) {
    return(0)
  } else {
    return(10)
  }
}
```

---

### Exiting a function

Most functions **return visibly**: calling the function prints the result:

```{r eval=TRUE}
j03 <- function() 1
j03()
```

--

Apply `invisible()` to the last value to prevent automatic printing:

```{r eval=TRUE}
j04 <- function() invisible(1)
j04()
```

---

### Exiting a function

**Print** or **wrap** function call in parentheses to make return value visible, or use `withVisible()`:

```{r eval=TRUE}
print(j04())

(j04())

str(withVisible(j04()))
```

---

### Exiting a function

**Errors**

If a function cannot complete its assigned task, it should **throw an error** with `stop()`, which immediately terminates the execution of the function.

```{r eval=FALSE}
j05 <- function() {
  stop("I'm an error")
  return(10)
}
j05()
#> Error in j05(): I'm an error
```

<br>

Learn more about error handling in **Chapter 8**

---

### Exiting a function

**Exit handlers**

Sometimes a function needs to make **temporary changes** to the **global state**. 

--

<br>
Use `on.exit()` to set up an **exit handler** with clean-up code that restores global state, even when functions exits with an error

--

```{r eval=FALSE}
cleanup <- function(dir, code) {
  old_dir <- setwd(dir)
  on.exit(setwd(old_dir), add = TRUE)
  
  old_opt <- options(stringsAsFactors = FALSE)
  on.exit(options(old_opt), add = TRUE)
}
```

<br>

**Note**: Always set `add = TRUE` to avoid overwriting previous exit handler

---

### Function forms

--

Function calls come in four varieties:

- **prefix**: function name before arguments `foofy(a,b,c)`

- **infix**: function name comes in between arguments: `x + y`

- **replacement**: function replaces values by assignment:<br>
`names(df)<-c("a","b","c")`

- **special**: functions like `[[`, `if` and `for`

--

<br>

You can always **rewrite** a function in **prefix form**:

```{r eval=FALSE}

x + y
`+`(x, y)

names(df) <- c("x", "y", "z")
`names<-`(df, c("x", "y", "z"))

for(i in 1:10) print(i)
`for`(i, 1:10, print(i))

```

---

### Function forms

Useful example with `lapply`:

```{r eval=TRUE}

add <- function(x, y) x + y
lapply(list(1:3, 4:5), add, 3)

```

--

<br>

You can get the same result with:

```{r eval=TRUE}

lapply(list(1:3, 4:5), `+`, 3)

```

---

### Function forms

**Infix functions**

Built-in examples: `:`, `::`, `$`, `@`, `^`, `*`, `/`, `+`, etc.

--

<br>

You can **create your own infix function**: bind two arguments to a name that starts and ends with `%`:

```{r eval=TRUE}

`%+%` <- function(a, b) paste0(a, b)
"new " %+% "string"

```

---

### Function forms

**Replacement functions**

Replacement functions act like they **modify their arguments in place**, and have the special name `xxx<-`

--

They must have arguments named `x` and `value`, and must return the modified object:


```{r eval=TRUE}
`second<-` <- function(x, value) {
  x[2] <- value
  x
}

```

--

Replacement functions are used by placing the function call on the left side of `<-`:

```{r eval=TRUE}
x <- 1:10
second(x) <- 5L
x
```
---

### Function forms

**Replacement functions**

Replacement functions **act like** they modify their arguments in place, and have the special name `xxx<-`

--

```{r eval=TRUE}
x <- 1:10
tracemem(x)
second(x) <- 6L
```

```{r include=FALSE}
untracemem(x)
```

---

### Function forms

**Replacement functions**

If your replacement function needs **additional arguments**, place them between `x` and `value`, and call the replacement function with additional arguments on the left:

```{r eval=TRUE}
`modify<-` <- function(x, position, value) {
  x[position] <- value
  x
}
modify(x, 1) <- 10
x
```

---

name: title
class: center, middle

## Thank you ##
 
Questions? Break for 10 min, and meet in your breakout rooms

_Check hackMD for your breakout room assignment_


