---
title: 'AdvancedR chapter 22: Debugging'
output:
  html_document:
    df_print: paged
---

```{r, echo = FALSE}

knitr::opts_chunk$set(fig.pos = "center")

library(dplyr)
library(ggplot2)
library(tidyr)

```

# Women in the Workforce (TidyTuesday 4-Mar-2019)

Data from the Bureau of Labor Statistics and the Census Bureau about women in the workforce. 

According to the AAUW - "The gender pay gap is the gap between what men and women are paid. Most commonly, it refers to the median annual pay of all women who work full time and year-round, compared to the pay of a similar cohort of men." 

The data is provided as is, and you recognize the limitations and issues in defining gender as binary.

```{r, message=FALSE, warning=FALSE}

jobs_gender <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/jobs_gender.csv")
head(jobs_gender)

```

```{r, out.width="140%", message=FALSE, warning=FALSE}

# Get mean_major category
mean_major_category <- jobs_gender %>% 
  group_by(major_category) %>%
  summarise(mean = mean(wage_percent_of_male/100, na.rm = T) )

# Get palette 
n_cols <- jobs_gender$minor_category %>% unique() %>% length()
color_array <- colorRampPalette(RColorBrewer::brewer.pal(8, "Set2"))(n_cols)

# Plot
jobs_gender %>%
  ggplot(aes(x = minor_category, y = wage_percent_of_male/100, color = minor_category)) +
  geom_boxplot(alpha = 0.3, show.legend = FALSE) +
  geom_point(position = position_jitter(width = 0.3)) +
  geom_hline(data = mean_major_category, aes(yintercept = mean), linetype = 2, color = "darkgray") +
  facet_wrap(~major_category, scales = "free_x") +
  scale_color_manual(values = color_array) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Wage of women in proportion to men in the same job category", y = "")+
  theme_light() +
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 8),
        axis.text.x = element_blank(),
        axis.title.x = element_blank())


mean <- mean_major_category$mean


```


## Debugging a for loop

Trying to store the average wage of women in proportion to men.

```{r, eval = FALSE}
# store the average wage percentage of males 
for(i in mean_major_category$major_category){
  prop <- jobs_gender[jobs_gender$major_category == i, "wage_percent_of_male"]
  proportions[i] <- mean(prop)
}
proportions
```

### Step 1: declaring a variable

```{r, eval = FALSE}
### debugging step by step

# problem 1: because 'changes' wasn't declared before the loop, it now refers to
# a dplyr function
proportions <- rep(NA, 8)

for(i in mean_major_category$major_category){
  prop <- jobs_gender[jobs_gender$major_category == i, "wage_percent_of_male"]
  proportions[i] <- mean(prop)
}

proportions
```

### Step 2: using a character variable for indexing

```{r, eval = FALSE}
proportions <- rep(NA, 8)
# problem 2: i is a character variable and cannot be used for indexing as is
# possible solution: assign names to changes
names(changes) <- mean_major_category$major_category
for(i in mean_major_category$major_category){
  prop <- jobs_gender[jobs_gender$major_category == i, "wage_percent_of_male"]
  # problem 1: changes wasn't declared before the loop
  proportions[i] <- mean(prop)
}

proportions
```
### Step 3: trying to use a tibble as a vector

```{r, eval = FALSE}
proportions <- rep(NA, 8)
names(proportions) <- mean_major_category$major_category

for(i in mean_major_category$major_category){
  prop <- jobs_gender[jobs_gender$major_category == i, "wage_percent_of_male"]
  # problem 3: change is a tibble and not a numeric or logical vector.
  proportions[i] <- mean(prop$wage_percent_of_male)
}

proportions
```
### Step 4: remove missing values in order to compute mean

```{r}
proportions <- rep(NA, 8)
names(proportions) <- mean_major_category$major_category

for(i in mean_major_category$major_category){
  prop <- jobs_gender[jobs_gender$major_category == i, "wage_percent_of_male"]
  # problem 4: missing values need to be removed in order to compute the mean. 
  proportions[i] <- mean(prop$wage_percent_of_male, na.rm = TRUE)
}

proportions
```

This is nonsensical code of course..
Maybe a good demonstration of how not to use a for loop! 
Perhaps this is a more realistic application: (currently unfinished)

```{r, eval = FALSE}
# store the average yearly change of the wage gap per major category
change <- jobs_gender %>%
  group_by(year, major_category) %>%
  summarise(mean_i <- mean(wage_percent_of_male, na.rm = TRUE))

for(i in unique(jobs_gender$year)){
  stopifnot(i < 2016)
  
  next_year <- i + 1 
  changes[i] <- change %>%
    group_by(major_category)
  
}
```


## Free ggplot2 exercise: how to plot group means?

```{r}
jobs_gender %>%
  group_by(major_category) %>%
  ggplot() + 
  geom_point(aes(x = mean(percent_female), 
                 y = mean(wage_percent_of_male), 
                 color = major_category)) + 
  theme_classic()

```

